---
title: "Missing data and the closing age in `ungroup`"
subtitle: "Behaviour before and after, for issues #4 and #6"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# Run against the dev tree. devtools::load_all() from the repository root works
# just as well if the package is not installed.
library(ungroup)
```

Two open issues ask for the same underlying thing: the ability to fit a
PCLM surface when parts of it were never observed.

* **#4** "NA's not allowed": the open age group moves over time, so a
  rectangular surface has unobserved cells in the top of the early years.
* **#6** "Allow for missing years": exposures are on a coarser time grid than
  deaths, so whole years of exposure are absent.

Both were hard errors. The fixes add two arguments, both appended after
`control` so every existing call keeps working:

| argument | default | meaning |
|---|---|---|
| `na.action` | `"fail"` | `"fail"` rejects `NA` exactly as before. `"omit"` drops those rows and lets the penalty bridge the gap. |
| `omega` | `NULL` | Closing age of the distribution. An alternative to `nlast`: `nlast` is taken as `omega - max(x)`. |

Only `NA` counts as unobserved. `Inf` stays an error under `"omit"`, so
"missing" and "invalid" remain distinguishable.

```{r surface}
x     <- c(0, 5, 10, 15, 20)   # five age groups; the last runs [20, 45)
nlast <- 25
years <- 1980:1989

base <- c(120, 65, 50, 45, 35)
y <- round(outer(base, seq(1, 0.8, length.out = length(years))))
colnames(y) <- years
y
```

---

## Issue #4: the open age group moves over time

The Bangladesh case in one line of history: 1980 to 1982 were closed at 10+,
1983 and 1984 at 15+, and only later years reach 20+. The surface has to be
rectangular, so the top left corner is simply unobserved.

```{r ragged}
y_ragged <- y
y_ragged[4:5, 1:3] <- NA    # 1980-82 closed at 10+
y_ragged[5,   4:5] <- NA    # 1983-84 closed at 15+
y_ragged
```

### Before

`na.action` defaults to `"fail"`, which is the old behaviour preserved
verbatim. This is what v1.4.5 did, and what a caller who never hears about
these arguments still gets.

```{r before4, error=TRUE}
pclm2D(x = x, y = y_ragged, nlast = nlast, verbose = FALSE)
```

The only options were to truncate every year to the smallest common open age
group, throwing away the rest of the data.

### After

```{r after4}
fit4 <- pclm2D(x = x, y = y_ragged, nlast = nlast,
               na.action = "omit",
               control = list(lambda = c(1, 1)),
               verbose = FALSE)

round(fitted(fit4), 1)
```

The unobserved cells come back as estimates rather than gaps. The fine grid is
untouched by the omission, so the shape of the output is exactly what a
complete surface would have produced.

```{r after4-plot, fig.width=7, fig.height=5}
plot(fit4, type = "fitted")
```

The whole surface is still one smooth estimate, because the age and year
penalties borrow strength across the gap rather than fitting the hole in
isolation.

### The closing age

Issue #4 also asked for an `omega`, a fixed closing age, instead of `nlast`.
The distinction matters when the last age group starts in different places:
`nlast` describes a width, `omega` describes where the life table ends.

```{r omega}
# These say the same thing from opposite ends.
a <- pclm(x = x, y = y[, 8], nlast = 25,
          control = list(lambda = 100), verbose = FALSE)
b <- pclm(x = x, y = y[, 8], omega = 45,
          control = list(lambda = 100), verbose = FALSE)

all.equal(unname(fitted(a)), unname(fitted(b)))
```

`omega` and `nlast` are alternatives, and the guard rails say so:

```{r omega-errors, error=TRUE}
pclm(x = x, y = y[, 8], nlast = 25, omega = 45)   # both
pclm(x = x, y = y[, 8])                           # neither
pclm(x = x, y = y[, 8], omega = 20)               # not past max(x)
```

---

## Issue #6: a year with no exposure

Deaths are observed every year. Exposures are not: here only every third year
was counted.

```{r exposures}
ex <- round(outer(c(3200, 3100, 3000, 2600, 2100),
                  seq(1, 1.05, length.out = length(years))))
ex[, setdiff(seq_along(years), seq(1, 10, by = 3))] <- NA
colnames(ex) <- years
ex
```

### Before

```{r before6, error=TRUE}
pclm2D(x = x, y = y, nlast = nlast, offset = ex, verbose = FALSE)
```

### After

The exposure is ungrouped on its own first pass, so it needs the same
treatment as the counts. `na.action` is carried into that pass.

```{r after6}
fit6 <- pclm2D(x = x, y = y, nlast = nlast, offset = ex,
               na.action = "omit",
               control = list(lambda = c(1, 1)),
               verbose = FALSE)

# Rates, so the display needs more than one decimal.
signif(fitted(fit6), 3)
```

Every year now has a rate, including the seven that were never counted. The
estimate at those years is interpolation from the neighbours, not a
resurrection of the underlying counts. `fitted()` returns rates here because an
offset was supplied.

```{r after6-plot, fig.width=7, fig.height=5}
plot(fit6, type = "fitted")
```

---

## Where omission and redistribution part company

Issue #4 proposed a different treatment for the ragged top: recode the
trailing `NA` series to zero and redistribute the last observed `Dx` and `Nx`
over `[last observed x, omega]`. Omission and redistribution agree everywhere
the data is dense and disagree in one specific place.

When a cohort dies out before reaching the closing age, redistribution forces
those deaths into the tail by construction. Omission leaves the tail to the
smoothing penalty, which is **extrapolation**, and it carries no coverage
guarantee.

That is what a smoother should do, and it is what the issue asked for in its
general form ("you can simply smooth over them"), but it is worth stating
plainly: the tail of an omitted fit is a model estimate with the weakest
support in the whole surface.

The confidence bands make no promise about that region either. `ci$conf_lower`
and `ci$conf_upper` are pointwise marginal intervals around the fit; where
nothing was observed they are as wide as the penalty allows and no wider.

```{r session}
sessionInfo()
```
