---
title: "`pclm` Installation"
author: "Marius D. Pascariu and Maciej J. Danko"
date: "December 8, 2017"
output: 
  pdf_document:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{pclm Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The package can be installed from two different sources, `CRAN` or `GitHub`. Most people would choose the first option (see below).

# Install stable varsion from `CRAN`

The `pclm` package can be installed from [`CRAN`](https://cran.r-project.org) as follows:

1. Make sure you have the most recent version of R
2. Run the following code in your R console 

```{r, eval=FALSE}
install.packages("pclm")
```

## For Linux users only

Note that `pclm` package depends on `rgl` in order to provide better graphics. Prior to installing `pclm` on a Linux operating system, the `rgl` package can be installed as follow:

```{r, eval=FALSE}
sudo apt-get install r-cran-rgl
```

You migth need to install manually other related packages:


```{r, eval=FALSE}
sudo apt-get install libfreetype6-dev
sudo apt-get install libftgl-dev
sudo apt-get install mesa-common-dev 
sudo apt-get install libglu1-mesa-dev
```

# Install development version from `GitHub`
The latest version can be installed from [`GitHub`](https://github.com/mpascariu/pclm) as follows:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("mpascariu/pclm")
```

In this case you will need to have `devtools` package already installed in order to allow R to connect to GitHub. Otherwise uncomment the fist line and install it.


